
\section{Proof of equivalence between MSE and cosine loss in predicting Sine and Cosine of $\chi$
angles}
Let ${\chi_i}$ and $\hat\chi_i$ be the true and the predicted angle values for a residue's $i^{th}$
$\chi$ angle, respectively. Let ${{\bf v}_i\equiv~[\sin {\chi_i},\cos {\chi_i}]}$ and ${\hat{\bf
v}_i\equiv[\sin { \hat \chi_i},\cos  {\hat \chi_i}]}$ denote their respective 2D vectors on the
unit circle, whose two components are the sine and cosine of the angles themselves. Then we have:
\begin{align}
 \nonumber    \text{MSE}_{\text{sin-cos}}(\hat\chi_i, \chi_i)
    \nonumber                                &= \frac{1}{2}(\cos \hat \chi_i - \cos \chi_i)^2 + \frac{1}{2}(\sin\hat \chi_i - \sin \chi_i)^2 \\
                                    % &= \frac{\cos(\hat{y}_{\chi})^2}{2} + \frac{\cos(y_{\chi})^2}{2} - \cos(\hat{y}_{\chi})\cos(y_{\chi}) + \frac{\sin(\hat{y}_{\chi})^2}{2} + \frac{\sin(y_{\chi})^2}{2} - \sin(x)\sin(y_{\chi}) \\
             \nonumber                       &= 1 - \cos \hat \chi_i  \cos {\chi_i} - \sin \hat \chi_i  \sin {\chi_i} \\
               \nonumber                     &= 1 - \langle \hat{\bf v}_{i},{\bf v}_i \rangle \\
                                    &= \text{CosineLoss}(\hat{\bf v}_{i}, {\bf v}_i)
    \label{eqn:proof_of_equivalence_sin_cos}
\end{align}
where the jump from step 1 to step 2 is granted by the trigonometric identity $\sin^2 \theta +
\cos^2 \theta = 1$.


\section{Training Details}
\label{sec:training_details}

\subsection{Toy Task}
All models were constructed with 5 equivariant blocks and no invariant feed-forward neural network
(FFNN) for the two models predicting invariant quantities. All models were trained using
Adam~\citep{kingma_adam_2017} for 30 epochs with learning rate 0.001 and a batch size of 32. The
model exhibiting lowest validation loss was used.

\subsection{Side-chain Packing}
Models with $\ell_{\text{max}} = 4$ have five equivariant blocks with a per-$\ell$ hidden feature
side of 128. For models with $\ell_{\text{max}} = 5$ we use 5 blocks with per-$\ell$ size 96. In
doing so, the two models have comparable number of parameters (3M). All models have a 3-layer FFNN
with silu nonlinearity and dropout normalization rate of 0.1. We found it useful to tune he dropout
rate to prevent overfitting. We train all models for 10 epochs, keeping the model with lowest
validation loss at the end of an epoch (convergence usually happened by epoch $\sim$8); models with
$\ell_{\text{max}} = 5$ took roughly 50 minutes per epoch to train on a single NVIDIA A40 GPU,
while models with $\ell_{\text{max}} = 4$ took 35 minutes.